---
title: "Week #2"
author: "Ryan Friedman"
format: html
editor: visual
---

## Vectors

- Make a simple vector, then check structure and typeof:

```{r}
x <- c(1, 2, 3, 4, sqrt(2))
x
str(x)
typeof(x)
```

- Make a vector of strings, then check structure and typeof:

```{r}
y <- c("fish", "chicken", "beef")
y
str(y)
typeof(y)
```

- Combine the vectors, then check structure and typeof:

```{r}
z <- c(x,y)
z
str(z)
typeof(z)
```

Note how the vector change (*coerces*) all elements to type "character"

- sort() will put results in order

```{r}
z
sort(z)
```

- Use seq() and its shortcuts to create vectors

```{r}
seq(from = 1, to = 6, by = 2) # from 1 to 6, counting by 2
seq(1, 6, 2) # same thing
seq(1, 4, length=5) # vector length 5, starting at 1 and ending at 4, evenly spaced
seq(0, 100, length = 6) # 0 to 100 in 6 steps
0:6 # shortcut for counting 0 to 6
0:6*20 # scaling the shortcut by multiplication
```

- Use runif() ("random uniform number generator") to make a vector of random numbers.

```{r}
set.seed(42) # sets a seed for random numbers to ensure execution of this code doesn't change each time
runif(4, 2, 3) # 4 random numbers, each between 2 and 3
runif(6) # 6 random numbers, each between 0 and 1
```

- Vectors can have attributes, such as names for the elements

```{r}
u <- c("a" = 1, "b" = 2, "c" = 3)
u
attributes(u)
names(u)
```

- Apply names

```{r}
names(x) = c("First", "Second", "Third", "Fourth")
x
```

- Accessing elements of a vector

```{r}
letters # a built-in vector of all 26 letters
letters[1] # the first element of the vector (counting start at 1)
letters[1:4] # elements 1 through 4
letters[u] # uses u=(1, 2, 3) from before as an indexing vector
letters[-5] # negatives remove elements
letters[-1*c(1, 2, 3, 5, 8, 12, 13)] # multiple negatives removing elements
u # remember u is a vector with names
u[c("a", "c")] # names can be used to access elements
```

- Factor vectors are created with factor() and are used when the elements are categorical data

```{r}
pets <- c("cat", "dog", "bird", "cat", "other", "bird", "bird", "other", "dog") # our data for pets as a list of categorical observations
pets_factor <- factor(pets, levels = c("cat", "dog", "bird", "rodent", "other")) # associates the defined categorical levels with the pets data
pets_factor
str(pets_factor) # reveals that factors store values as integers, each associated with a level
```

- Levels need to align with the values in the data but we can use Labels to have different values show

```{r}
pets_factor <- factor(pets,
               levels = c("cat", "dog", "bird", "rodent", "other"),
               labels = c("Meow", "Woof", "Tweet", "Sniff", "zzzzz"))
pets_factor
```

- as.factor() is a shortcut to convert to factor vector, but only uses the levels visible

```{r}
pets_factor <- as.factor(pets) # makes factor vector from all levels observed in pets
pets_factor # notice rodent gets skipped
levels(pets_factor) <- c(levels(pets_factor), "rodent") # here rodent is manually added in afterwards
pets_factor
```

## Matrices

- Matrices are homogenous, containing vectors of the same type and length. Each column of the matrix is a vector. We can also think of matrix as a long vector we give dimensions to

- Create a matrix

```{r}
matrix(data = NA, 
       nrow = 1, 
       ncol = 1, 
       byrow = FALSE,
       dimnames = NULL) # syntax
my_mat <- matrix (c(1,3,4,-1,5,6),
                  nrow = 3,)  # If only nrow is provided, ncol is inferred
my_mat # note how data fills the colummns one at a time, but this can be changed with byrow.
```

- Coerce the matrix back into a vector

```{r}
as.vector(my_mat)
```

- If we under-fill a matrix with not enough data, we'll get a warning and the data repeats

```{r}
matrix(c(x, y, z), ncol = 3)
```

- Subset a matrix from dataframe Iris

```{r}
my_iris <- as.matrix(iris[, 1:4]) # takes all rows, and columns 1:4 of Iris and converts to a matrix
head(my_iris) # shows the beginning rows of the new matrix
```

- Manually reassigning names to the rows and columns using list()

```{r}
dimnames(my_iris) <- list(
  1:150, #first list element is a vector for the row names
  c("Sepal Length", "Sepal Width", "Petal Length", "Petal Width") #second element is a vector for column names
)
head(my_iris)
```

- We can also assign names when a matrix is created

```{r}
my_mat3 <- matrix(c(runif(10), 
                    rnorm(10),
                    rgamma(10, shape = 1, scale = 1)),
                  ncol = 3,
                  dimnames = list(1:10, c("Uniform", "Normal", "Gamma")))
my_mat3
```

- Accessing elements of a matrix using \[\]

```{r}
mat <- matrix(c(1:4, 20:17), ncol = 2)
mat
mat[2, 2]
```

- Arrays are n-dimensional matrices

```{r}
my_array <- array(1:24, dim = c(4, 2, 3))
my_array
```

## Data Frames

- A data frame is a list of vectors of the same length. The vectors can be different types.
- Create a data frame using the data.frame() function

```{r}
x <- c("a", "b", "c", "d", "e", "f")
y <- c(1, 3, 4, -1, 5, 6)
z <- 10:15
my_df <- data.frame(x, y, z) # combines three equal length vectors of different types
my_df
```

- Data frames take on the column names of the variables we use, but we can also set the names as we create the dataframe

```{r}
my_df <- data.frame(char = x, data1 = y, data2 = z)
my_df
```

- We can access elements of a data frame just like a matrix

```{r}
iris[1:4, 2:4] #returns a data frame
```

- If we access a single column, it can return a vector or a data frame, depending on whether drop=FALSE is used

```{r}
iris[1:10, 1] #returns a vector
```

```{r}
iris[1:10, 1, drop = FALSE] #return a data frame
```

- We can also call the \[ function as a prefix function like this:

```{r}
`[`(iris, 1:10, 1, drop = FALSE)
```

- Column names can be used for subsetting in place of the column numbers

```{r}
iris[1:5 , c("Sepal.Length", "Species")] # will just select those columns named
```

- Use \$ to access a single column, or do it with \[\]

```{r}
iris$Sepal.Length
iris[,"Sepal.Length"]
```

- Check names of the columns using colnames()

```{r}
colnames(iris)
```

## Lists

- Use list() to construct from any objects of any type or size

```{r}
my_list <- list(My_Data_Frame = my_df, Norm_Vals = rnorm(4), Punctuation = c("!", "?"))
my_list
str(my_list)
```

- Access list elements as a sub-list using \[\]. \[\[\]\] will return the element not as a list

```{r}
my_list[2]
my_list[[2]]
my_list[[2]][1:3] # selects the second item of the list as a vector, then shows 1:3 of that vector
```

- Note: Technically a data frame is considered a list

```{r}
typeof(my_df)
is.data.frame(my_df)
is.list(my_df)
```

## Control Flow: Logicals & if/then/else

- Logical Statements for comparison will return TRUE or FALSE

```{r}
"hi" == "hi"
"Hi" == "hi" # string equivalence must be exact
4 >= 3
3 != 4
TRUE == 1 # TRUE and 1 are equivalent
sqrt(3)^2 == 3 # returns FALSE due to loss of precision
dplyr::near(sqrt(3)^2, 3) # uses near() function from the dplyr package
```

- is. functions for verifying identity or properties

```{r}
is.numeric("ABC")
is.numeric(c(1, 2,3 ))
is.character(c("a","130","word"))
is.na(c(1, 2, 3, NA, 5, NA, 7)) # note is.na( ) will evaluate each element
```

- Using logic to subset data

```{r}
iris[iris$Species == "setosa", ] # will subset the rows of iris where species is "setosa"
```

- Compound logic: & for AND, \| for OR

```{r}
iris[iris$Petal.Length > 1.5 & iris$species == "setosa", ] # subset rows with long petals and setosa
iris[((iris$Petal.Length > 1.5) | (iris$Petal.Width < 0.15)) & (iris$Species == "setosa"), ] # subset rows that are setosa and meet a particular length or width requirement
```

- if/then/else syntax

```{r}
head(airquality)
if(airquality$Wind[1] >= 15) { 
  "High Wind"
} else if (airquality$Wind[1] >= 10){
  "Windy"
} else if (airquality$Wind[1] >= 6) {
  "Light Wind"
} else if (airquality$Wind[1] >= 0) {
  "Calm"
} else {
  "Error"
}
```

Fizz Buzz challenge

```{r}
number <- 4
if((number %% 15) == 0){
  print("fizz buzz")
} else if ((number %% 5) == 0){
  print("buzz")
} else if ((number %% 3) == 0){
  print("fizz")
} else {
  print("whoops?")
}
```

## Control Flow: Loops

- for() loop

```{r}
for (i in 1:10){
  print(i)
}
```

- The index can be a vector rather than a counting number

```{r}
for (index in c("cat","hat","worm")){
  print(index)
}
```

- create vectors of length 5, then use a loop to print 5 sentences

```{r}
words<-c("first", "second", "third", "fourth", "fifth")
data <- runif(5) # 5 random numbers
for (i in 1:length(data)){
  print(paste0("The ", words[i], " number is ", data[i]))
}
```

- use break to escape a loop, or next to skip an index

- Apply loops with Fizz Buzz

```{r}
for(number in -1:41){
  if((number %% 15) == 0){
    print(paste0(number, " fizz buzz"))
  } else if ((number %% 5) == 0){
    print(paste0(number, " buzz"))
  } else if ((number %% 3) == 0){
    print(paste0(number, " fizz"))
  } else {
    print(paste0(number, " whoops?"))
  }  
}
```

## Control Flow: Vectorized Functions

- Set up my_batting data frame

```{r}
library(Lahman)
my_batting <- Batting[, c("playerID", "teamID", "G", "AB", "R", "H", "X2B", "X3B", "HR")]
head(my_batting)
```

- use colMeans() to find the mean for an entire column

```{r}
colMeans(my_batting[, 3:9])
```

- we can also do colMedians(), but we need to convert to matrix

```{r}
library(matrixStats)
colMedians(as.matrix(my_batting[, 3:9]))
```

- Alternatively this could be done by looping through all the columns and calculating each mean. Let's try both and check the computation speed. Looping is slower

```{r}
library(microbenchmark)
my_numeric_batting <- Batting[, 6:22] #get all numeric columns
vectorized_results <- microbenchmark( 
  colMeans(my_numeric_batting, na.rm = TRUE) # uses colMeans
)

loop_results <- microbenchmark(
  for(i in 1:17){
    mean(my_numeric_batting[, i], na.rm = TRUE) # loops through each column
  }
)
vectorized_results
loop_results
```

- Let's try ifelse(vector_condition, if_true_do_this, if_false_do_this)

```{r}
ifelse(airquality$Wind >= 15, "HighWind", "Not HighWind")
```

- Also can do nested ifelse

```{r}
ifelse(airquality$Wind >= 15, "HighWind",
          ifelse(airquality$Wind >= 10, "Windy",
                 ifelse(airquality$Wind >= 6, "LightWind", 
                        ifelse(airquality$Wind >= 0, "Calm", "Error"))))
```

### Summary: Vetorized functions are faster than looping

## Writing Functions

- Lets create a standardize() function that will input a vector of values and then convert each into z-scores (based on the vector's mean and stdev)

```{r}
standardize <- function (vector){
  return((vector-mean(vector)) / sd(vector))
}
```

- Create data and apply the function

```{r}
set.seed(10)
data <- runif(15)
data
result <- standardize(data)
result
mean(result)
sd(result)
```

- Now add options for centering and scaling. They are TRUE by default

```{r}
standardize <- function (vector, center = TRUE, scale = TRUE){
  if (center){
    vector <- vector-mean(vector)
  }
  if (scale){
    vector <- vector / sd(vector)
  }
  return(vector)
}
data
print(standardize(data,FALSE,FALSE))
print(standardize(data,TRUE,FALSE))
print(standardize(data))
```

- Now let's also return mean and sd of the data

```{r}
standardize <- function (vector, center = TRUE, scale = TRUE){
  mean <- mean(vector)
  stdev <- sd(vector)
  if (center){
    vector <- vector-mean
  }
  if (scale){
    vector <- vector / stdev
  }
  return(list(result = vector, mean = mean, sd = stdev)) # adding names to the list items
}
print(standardize(data))
```

- Use stop() to exit a function. Use switch() as an alternative to if/then/else to pick between options. Create summarizer() function, which performs requested calculations but also warns if input type is incorrect

```{r}
summarizer <- function(vec, type, trim = 0.05) {
  if(!is.vector(vec) | !is.numeric(vec)){
    stop("Not a vector or not numeric my friend.")
  }
  switch(type, 
         mean = mean(vec),
         median = median(vec),
         trimmed = mean(vec, trim),
         stop("Mistake!")
         )
}
```

- Execute summarizer function

```{r}
summarizer(letters, "mean")
```

```{r}
summarizer(c(1,1,1,6,10), "mean")
```

```{r}
summarizer(c(1,1,1,6,10), "trimmed", 0.2)
```

```{r}
summarizer(c(1,1,1,6,10), "mean")
```

- For arguments in a function, we can choose whether or not to name them. R uses positional matching for all inputs not explicitly named.

- An infix function is a function that goes between arguments (as opposed to a prefix function that goes before arguments). Here's an example "+" used as both an infix and a prefix function

```{r}
3 + 4
'+'(3, 4)
```

Creating our own infix function %+% for pasting strings

```{r}
'%+%' <- function(a,b){
  paste0(a, b)
}
"Test " %+% "pasting " %+% "strings."
```

- Base R Pipe is used as a shorthand for nested functions/parenthesis. The two code sections are identical:

```{r}
library(dplyr)
arrange(select(filter(as_tibble(Lahman::Batting), teamID == "PIT"), playerID, G, X2B), desc(X2B))
```

```{r}
Lahman::Batting |> #read the pipe as "then"
  as_tibble() |>
  filter(teamID == "PIT") |>
  select(playerID, G, X2B) |> 
  arrange(desc(X2B)) 
```
