---
title: "Week #4"
format: pdf
editor: visual
---

## Querying APIs

- API:

  - Application Programming Interface

  - A protocol for computers to talk to each other, allowing us to grab data from systems.

  - Also we can set up APIs to allow others access to our data

  - Many websites have require an API key

- The API for the USA Census (census.gov) is accessible through a package `tidycensus`j

```{r}
library(tidycensus)
```

- Try `get_acs()` to obtain data from the American Community Survey

```{r}
rent <- "DP04_0142PE" # PE means percentage
rent_data <- get_acs(variables = rent,
                     geography = "county",
                     geometry = TRUE, # returns the polygon data and allows for maps easily
                     survey = "acs5",
                     show_call = TRUE, # can add state and other things
                     key = "5a644b1ecc301cab1403e0940017740912735ef9") 
```

- Install and try `mapview`

```{r}
library(mapview)
library(leaflet)
library(leafem)
library(sf)
```

```{r}
rent_data |>
  mapview(zcol = "estimate",
          layer.name = "Median rent as a % of gross income")
```

- JSON data (Java Script Object Notation)

- Recommended package: `jsonlite`

```{r}
library(jsonlite)
```

```{r}
library(httr)
URL_ids <- "https://api.nhle.com/stats/rest/en/team"
id_info <- GET(URL_ids)
str(id_info, max.level = 1)
```

- status "200" means it worked

- the thing we are interested in is content, which is json data

```{r}
library(tidyr)
parsed <- fromJSON(rawToChar(id_info$content))
team_info <- as_tibble(parsed$data)
team_info 
```

## EDA Concepts

- Basic steps of Exploratory Data Analysis (EDA)

  - Understand how data is stored

  - Basic data validation

  - Determine rate of mising values

  - Clean up data as needed

  - Investigate distributions (univariate, multivariate)

  - Apply transformations

```{r}
library(tidyverse)
library(readxl)
app_data <- read_excel("PracticeDataFiles/app_data.xlsx")
app_data
```

- note: BMI read in as a chr, though it should be a dbl. Can use as.numeric to fix
- Also US_number would be better off as a char since it is categorical

```{r}
app_data <- app_data |>
  mutate(BMI = as.numeric(BMI),
         US_Number = as.character(US_Number))
app_data
```

- use describe() to get basic numeric summary of all

```{r}
library(psych)
describe(app_data)
```

- use is.na() to determine if any NA values exist

```{r}
colSums(is.na(app_data))
```

- drop_na() will remove rows that have NA

- replace_na() will replace NA observations with a specific value

- Factor is great to apply to categorical data, as long as all of the levels are known

- use unique() to identify possible factor levels

```{r}
unique(app_data$Sex)
unique(app_data$Diagnosis)
unique(app_data$Severity)
```

Make new factor variables for these:
```{r}
app_data <- app_data |>
  mutate(SexF = factor(Sex, levels = c("female", "male"), labels = c("Female", "Male")),
         DiagnosisF = as.factor(Diagnosis),
         SeverityF = as.factor(Severity))
app_data |>
  select(SexF, DiagnosisF, SeverityF)
```

- Contingency Tables using summarize()

- One-way table
```{r}
app_data |>
    group_by(SexF) |> # summarize respects the groupings
    drop_na(SexF) |>
    summarize(count = n()) # defines new variable 'count', while n() just adds up the qty
```

- Two-way table
```{r}
app_data |>
    group_by(SexF, DiagnosisF) |> 
    drop_na(SexF, DiagnosisF) |>
    summarize(count = n()) |> # defines new variable 'count', while n() just adds up the qty
    pivot_wider(names_from = DiagnosisF, values_from = count)
```

- Bar Charts for visualization of categorical data

- Getting mean. median, standard deviation, etc using summarize()

```{r}
app_data |>
    group_by(Diagnosis) |>
    drop_na(Diagnosis, Weight) |>
    summarize(mean_weight = mean(Weight),
              median_weight = median(Weight),
              stdev_weight = sd(Weight))
```

More flexible usage of summary() with across() and auto generated names

```{r}
app_data |>
  group_by(Diagnosis) |>
  drop_na(Diagnosis, Weight) |>
  summarize(across(Weight, .fns = list ("mean" = mean,
                                        "median" = median,
                                        "var" = var,
                                        "sd" = sd,
                                        "IQR" = IQR), .names = "{.fn}_{.col}"))
  
  
```

- covariance and correlation

```{r}
app_data |>
  drop_na(Weight, Age) |>
  summarize(cov = cov(Weight, Age), cor(Weight, Age))
```

- again finding the above, now by group

```{r}
app_data |>
  drop_na(Weight, Age) |>
  group_by(Diagnosis) |>
  summarize(cov = cov(Weight, Age), cor(Weight, Age))
```

