---
title: "Week #9"
format: pdf
editor: visual
---

```{r}
library(tidyverse)
```

## Modeling Concepts

```{r}
# read in bike data and create some variables
bike_data <- read_csv("https://www4.stat.ncsu.edu/~online/datasets/bikeDetails.csv")
bike_data <- bike_data |> 
  mutate(log_selling_price = log(selling_price), 
         log_km_driven = log(km_driven)) |>
  select(log_km_driven, log_selling_price, everything())
bike_data
# log selling price is the target(response) variable
```

- perform a regression task, i.e. predicting the response values

  - log_selling_price is the response variable

  - log_km_driven is the predictor variable

```{r}
ggplot(bike_data, aes(x = log_km_driven, y = log_selling_price)) +
  geom_point() +
  geom_smooth(method = "lm") # fits simple linear regression line
```

```{r}
ggplot(bike_data, aes(x = log_km_driven, y = log_selling_price)) +
  geom_point() +
  stat_smooth(method = "lm", formula = y ~ x + I(x^2))
# quadratic predictor (i.e. linear model, but the predictor is squared)
```

```{r}
ggplot(bike_data, aes(x = log_km_driven, y = log_selling_price)) +
  geom_point() +
  stat_smooth(method = "lm", formula = y ~ x + I(x^2) + I(x^3))
# here we have an x^3 term
```

- Need to quantify how well the model is predicting

```{r}
ggplot(bike_data, aes(x = log_km_driven, y = log_selling_price)) +
  geom_point() +
  geom_smooth()
# generalized additive model
```

- Also consider regression tree models (disjointed non-linear)

- Loss function: criteria used to fit a model (i.e. y-yhat, (y-yhat)\^2, or \|y-yhat\|)

- Goal is to optimize/minimize the loss function

- `lm()` is the linear model function

- use formula notation: `y ~ x -> response ~ model terms`

```{r}
slr_fit <- lm(log_selling_price ~ log_km_driven, data = bike_data)
slr_fit
```

- use summary() on the model to get p-values for the coefficients

```{r}
summary(slr_fit)
```

- use plot() on the model for fitted diagnostics

```{r}
plot(slr_fit)
```

- use `predict()` to test our fit with new data

```{r}
predict(slr_fit, newdata = data.frame(log_km_driven = c(log(1000), log(10000), log(100000))))

```

```{r}
# use exp to exponentiate (reverse) the log
exp(predict(slr_fit, newdata = data.frame(log_km_driven = c(log(1000), log(10000), log(100000)))))
```

- we use a [loss]{.underline} function to fit the model, and a [metric]{.underline} to evaluate the model

- metric can be the same as the loss, but not necessarily

- often we use RMSE (root mean squared error) as our metric

- Alternatives:  MSE (mean squared error) or MAE (mean absolute error) or Huber Loss (doesn't penalize large misakes as much)

```{r}
# predictions for the x's the model was trained on
head(predict(slr_fit))
```

```{r}
# MSE
mean((bike_data$log_selling_price-predict(slr_fit))^2)
# RMSE
sqrt(mean((bike_data$log_selling_price-predict(slr_fit))^2))
```

- Ideally train the model on a training data set, and then test the model on a test data set.