---
title: "Week #3"
format: pdf
editor: visual
---

## Packages

- Everything so far has been using Base R, which are the default packages that open when RStudio loads
- Packages, aka libraries or modules
- Collection of functions or datasets
- CRAN has the 'official' packages for R
- Environment Tab -\> Global environment shows packages loaded
- First time using a package on a computer you must install it using install.packages(). Can browse on Packages Tab

```{r}
# install.packages("dplyr") # Only needs to be done once
```

- Then you load the package using library(), which must be done each session

```{r}
# library("dplyr")
```

- To list everything in a package use ls():

```{r}
# ls("package:dplyr")
```

- To ensure a function is used from a specific package (in the case the function name exists in multiple packages), use ::

```{r}
# dplyr::filter(iris, Species == "virginica")
```

## Tidyverse Essentials

- Tidyverse is a collection of packages meant to work together that have similar syntax
- Tidyverse introduces a special kind of data frame called a **tibble**

```{r}
iris_tbl <- dplyr::as_tibble(iris)
class(iris_tbl)
```

- Class includes a data frame, but also other things

```{r}
str(iris_tbl)
```

- Structure looks a lot like data frame

```{r}
iris_tbl
```

- Default printing of the tibble is more organized

#### Tidyverse Packages:

- **ggplot2** is a system for declaratively creating graphics, based on The Grammar of Graphics. You provide the data, tell ggplot2 how to map variables to aesthetics, what graphical primitives to use, and it takes care of the details

- **dplyr** provides a grammar of data manipulation, providing a consistent set of verbs that solve the most common data manipulation challenges

- **tidyr** provides a set of functions that help you get to tidy data. Tidy data is data with a consistent form: in brief, every variable goes in a column, and every column is a variable

- **readr** provides a fast and friendly way to read rectangular data (like csv, tsv, and fwf). It is designed to flexibly parse many types of data found in the wild, while still cleanly failing when data unexpectedly changes

- **purrr** enhances R’s functional programming (FP) toolkit by providing a complete and consistent set of tools for working with functions and vectors. Once you master the basic concepts, purrr allows you to replace many for loops with code that is easier to write and more expressive

- **tibble** is a modern re-imagining of the data frame, keeping what time has proven to be effective, and throwing out what it has not. Tibbles are data.frames that are lazy and surly: they do less and complain more forcing you to confront problems earlier, typically leading to cleaner, more expressive code

- **stringr** provides a cohesive set of functions designed to make working with strings as easy as possible. It is built on top of stringi, which uses the ICU C library to provide fast, correct implementations of common string manipulations

- **forcats** provides a suite of useful tools that solve common problems with factors. R uses factors to handle categorical variables, variables that have a fixed and known set of possible values

## Reading Delimited Data

- Delimited, meaning data separated by a specific character like ',' or ' '

- Load the tidyverse library into current session

```{r}
library(tidyverse)
```

- read_csv() requires a file argument, which can be a path to file, a connection, or literal data

- Example reading a csv file from the web

```{r}
bike_details <- read_csv("https://www4.stat.ncsu.edu/~online/datasets/bikeDetails.csv")
```

```{r}
bike_details
```

- Note columns come with names and column types, which should be confirmed

- When we subset a single column of a tibble, the result is a tibble. If we want the column as a vector, we can use \$ or pull() from dplyr.

```{r}
bike_details[1:10,"name"]
```

```{r}
bike_details$name[1:10]
```

```{r}
bike_details[1:10, ] |>
  pull(name)
```

- read_csv() is for comma separated files that have comma as the delimeter. There is also read_tsv() for tabs, read_table() for spaces, read_csv2() for semi colons, read_fwf() for fixed width fields.

- for other cases, uses read_delim(). Delimeter should be specified. Also if column names are not present in the file, set col_names = FALSE or col_names = c("First", "Second", "Third", ...)

- Test with another data file

```{r}
ump_data <- read_delim("https://www4.stat.ncsu.edu/~online/datasets/umps2012.txt", 
                       delim = ">",
                       col_names = c("Year", "Month", "Day", "Home", "Away", "HPUmpire")
)
```

```{r}
ump_data
```

- For fixing dates, we will want to use the lubridate package and ymd() function

- First create a column to store dates in, and give it some values

```{r}
ump_data$date <- ymd("2012-01-01")
ump_data
```

- Now loop through all rows and construct date from the data with ymd() and data from the other columns

```{r}
for (i in 1:nrow(ump_data)){
  ump_data$date[i] <- ymd(paste(ump_data$Year[i],
                                ump_data$Month[i],
                                ump_data$Day[i], 
                                sep = "-"))
}
ump_data
```

## Reading Excel Data

- we use the readxl package, which is not a default part of tidyverse so we need to load it

```{r}
library(readxl)
```

- Excel data has to be saved locally (no url). I'm putting it in a folder called PracticeDataFiles

```{r}
dry_bean_data <- read_excel("PracticeDataFiles/Dry_Bean_Dataset.xlsx")
dry_bean_data
```

- To look at specific sheets, use excel_sheets

```{r}
excel_sheets("PracticeDataFiles/Dry_Bean_Dataset.xlsx")
```

- Lets grab the second sheet

```{r}
citation_dry_bean_data <- read_excel("PracticeDataFiles/Dry_Bean_Dataset.xlsx", 
                            sheet = excel_sheets("PracticeDataFiles/Dry_Bean_Dataset.xlsx")[2])
citation_dry_bean_data
```

- To read specific cells in a contiguous range:

```{r}
dry_bean_range <- read_excel("PracticeDataFiles/Dry_Bean_Dataset.xlsx", 
                   range = cell_cols("A:B")
                   )
dry_bean_range
```

## Manipulating Data with dplyr

- Row Manipulations:

  - **Filter** i.e. subsetting

  - **Arranging** i.e. putting in order

- Column Manipulations:

  - **Selecting** i.e. grab a subset of variables

  - **Rename** a column

  - **Mutating** to create a new variable

- Example of chaining manipulations with \|\>:

```{r}
my_data |>                                
    select(var1:var10, var20) |>   
    filter(var1 == "first") |>   
    mutate(new_var = lag(var20)) |>   
    ...
```

- use as_tibble() to convert data frame to tibble if needed

- filter() takes a tibble as first argument and logical vector as second argument. Can do with or without chaining:

```{r}
filter(batting_tbl, G > 50)
# or
batting_tbl |>
  filter(G > 50)
```

- you have multiple logic conditions, you can use & or you can separate by comma:

```{r}
batting_tbl |>
  filter(G > 50 & yearID == 2018)

batting_tbl |>
  filter(G > 50, yearID == 2018)
```

- use %in% to chose any observations matching an element in a vector

```{r}
batting_tbl |>
  filter(G > 50, yearID %in% c(2018, 2019, 2020))
```

- arrange() is used to reorder rows. example including chaining will arrange first by descending teamID and then ascending playerID

```{r}
batting_tbl |>
  filter(G > 100, yearID %in% c(2018, 2019, 2020)) |>
  arrange(desc(teamID), playerID)
```

- To return a single columns we have options `dplyr::pull()` `$` and `[ , ]`

- To return multiple columns in the order desired, use select()

```{r}
batting_tbl |>
  filter(G > 100, yearID %in% c(2018, 2019, 2020)) |>
  arrange(desc(teamID), playerID) |>
  select(playerID, teamID, H, X2B, X3B, HR)
```

- This can also select a contiguous range of columns. In this case, H to HR

```{r}
batting_tbl |>
  filter(G > 100, yearID %in% c(2018, 2019, 2020)) |>
  arrange(desc(teamID), playerID) |>
  select(playerID, teamID, H:HR)
```

- Can also use starts_with() and ends_with()

```{r}
batting_tbl |>
  filter(G > 100, yearID %in% c(2018, 2019, 2020)) |>
  arrange(desc(teamID), playerID) |>
  select(ends_with("ID") | starts_with("X"), G, AB, H, HR)
```

- If we're just reordering, everything() is useful

```{r}
batting_tbl |>
  filter(G > 100, yearID %in% c(2018, 2019, 2020)) |>
  arrange(desc(teamID), playerID) |>
  select(playerID, H:HR, everything())
```

- rename() used to change the names of columns

```{r}
batting_tbl |>
  filter(G > 100, yearID %in% c(2018, 2019, 2020)) |>
  arrange(desc(teamID), playerID) |>
  select(playerID, teamID, H:HR) |>
  rename("Doubles" = "X2B", "Triples" = "X3B")
```

- mutate() is used to create new variables (also modify or delete)

```{r}
batting_tbl |>
  filter(G > 100, yearID %in% c(2018, 2019, 2020)) |>
  arrange(desc(teamID), playerID) |>
  select(playerID, teamID, H:HR) |>
  rename("Doubles" = "X2B", "Triples" = "X3B") |>
  mutate(Extra_Base_Hits = Doubles + Triples + HR,
         Singles = H - Extra_Base_Hits) |>
  select(playerID, teamID, Singles, Doubles:HR, H, Extra_Base_Hits) # select just used here to reorder columns
```

- There are also calculation functions you can use within mutate(). A couple examples here:

```{r}
batting_tbl |>
  filter(G > 100, yearID %in% c(2018, 2019, 2020)) |>
  arrange(desc(teamID), playerID) |>
  select(playerID, teamID, H:HR) |>
  rename("Doubles" = "X2B", "Triples" = "X3B") |>
  mutate(Extra_Base_Hits = Doubles + Triples + HR,
         Singles = H - Extra_Base_Hits,
         H_Percentile = percent_rank(H),
         H_Mean = mean(H)) |>
  select(playerID, teamID, H, H_Mean, H_Percentile, everything()) 
```

- Use group_by() if we want to apply the calculation just within the groups. Here the mean and percentile are calculated for each observation, but on a by team basis:

```{r}
batting_tbl |>
  filter(G > 100, yearID %in% c(2018, 2019, 2020)) |>
  arrange(desc(teamID), playerID) |>
  select(playerID, teamID, H:HR) |>
  rename("Doubles" = "X2B", "Triples" = "X3B") |>
  group_by(teamID) |>
  mutate(Extra_Base_Hits = Doubles + Triples + HR,
         Singles = H - Extra_Base_Hits,
         H_Percentile = percent_rank(H),
         H_Mean = mean(H)) |>
  select(playerID, teamID, H, H_Mean, H_Percentile, everything()) |>
  print(n = 50)
```

- If you want to get fancy, group_by() can take multiple arguments. `group_by(teamID, yearID)` would do the calculations by both year and team.

- ifelse() can be used to create new variables where the value is dependent upon conditions

```{r}
batting_tbl |>
  filter(G > 100, yearID %in% c(2018, 2019, 2020)) |>
  arrange(desc(teamID), playerID) |>
  select(playerID, yearID, teamID, H:HR) |>
  rename("Doubles" = "X2B", "Triples" = "X3B") |>
  group_by(teamID, yearID) |>
  mutate(Extra_Base_Hits = Doubles + Triples + HR,
         Singles = H - Extra_Base_Hits,
         H_Percentile = percent_rank(H),
         H_Mean = mean(H),
         Status = ifelse(H > H_Mean, 
                         "Great", 
                         "Needs some work")) |>
  select(playerID, teamID, yearID, H, H_Mean, Status, H_Percentile, everything())
```

- try clean_names() if the imported column names get funky

## Manipulating Data with tidyr

- `pivot_longer()` lengthens data by increasing the number of rows and decreasing the number of columns

- `pivot_wider()` - widens data by increasing the number of columns and decreasing the number of rows

- Separate a column using `separate_wider_delim()`, i.e. with dates

- Combine two columns with `unite()`

## Databases and Basic SQL

- A database is a group of tables that are related by keys. Some variables are common across multiple tables, and help relate the tables to each other. (i.e. relational database)

  - Relational Database Management System (RDBMS) examples

  - Oracle

  - MySQL

  - SQL Server

  - PostgreSQL

  - SQLite

  - Azure SQL

- Most RDBMS have their own Structured Query Language (SQL) for getting information from tables in the database

- Actions on databases (CRUD)

  - Create data (adding rows or a table to the DB)
  
  - Read data (return data from DB table)
  
  - Update data
  
  - Delete data
  
  - Also: access control, monitoring, backup/recovery, tuning
  
- `dbConnect()` is used to connect to a DB.  Needs the `DBI` package

```{r}
library(DBI)
con <- dbConnect(data_base_type_goes_here_usually_requires_a_package, 
  host = "hostname.website",
  user = "username",
  password = rstudioapi::askForPassword("DB password")
)
```

- Database types and their packages:

  - RSQLite::SQLite() for RSQLite
  
  - RMySQL::MySQL() for RMySQL
  
  - RPostgreSQL::PostgreSQL() for RPostgreSQL
  
  - odbc::odbc() for Open Database Connectivity
  
  - bigrquery::bigquery() for google’s bigQuery

- Once connected, use tbl() to reference a table in the database.

```{r}
new_data <- tbl(con, "name_of_table")
```

- When done, it's good practice to disconnect

```{r}
dbDisconnect(con)
```

#### DB Example

- Connect to lahman.db

```{r}
library(DBI)
con <- dbConnect(RSQLite::SQLite(), "PracticeDataFiles/lahman.db")
```

- List out the tables in the DB

```{r}
dbListTables(con)
```
- Make sure dplyr is loaded, and lets look at the pitching table:

```{r}
library(dplyr)
tbl(con, "Pitching")
```

- Note the number of rows is missing, because it doesn't pull all of the data until you ask with collect() or store the result in an object

```{r}
tbl(con, "Pitching") |>
  select(ends_with("ID")) |>
  filter(yearID == 2010) |>
  collect()
```

- We can see the SQL code in the background with `show_query()`

```{r}
tbl(con, "Pitching") |>
  select(ends_with("ID")) |>
  filter(yearID == 2010) |>
  show_query()
```

- You can also just write SQL code:

```{r}
tbl(con, sql(
"SELECT `playerID`, `yearID`, `teamID`, `lgID`
FROM `Pitching`
WHERE (`yearID` = 2010.0)")
)
```

- See Video in course content on how to connect to Google BigQuery