---
title: "Week #6"
format: pdf
editor: visual
---

## apply() family of functions

- apply() family of functions are faster than for loops
- apply() will apply a function over an array or matrix
- lapply() will apply a function to every element of a list
- sapply() is a simplified version

```{r}
library(Lahman)
library(dplyr)
library(tidyr)
```

```{r}
my_batting <- Batting[, c("playerID", "teamID", "G", "AB", "R", "H", "X2B", "X3B", "HR")] |>
  as_tibble()
my_batting
```

- Use apply() to find summaries of the columns

```{r}
apply(X = my_batting,
  MARGIN = 2, # apply to each column
  FUN = summary,
  na.rm = TRUE)
```

- Note: Because apply() assumes similar data types, summary gives us unhelpful results since everything is coerced to characters.

- Lets try again, this time just selecting numeric data

```{r}
apply(X = my_batting |>
        select(where(is.numeric)),
  MARGIN = 2, # apply to each column
  FUN = summary,
  na.rm = TRUE)
```

- Now we get useful results

- Sometimes we can define a temporary function with apply(). This is known as an *anonymous* function or *lambda* function

```{r}
custom_batting_summary <- apply(X = my_batting |>
                                  select(where(is.numeric)),
                                MARGIN = 2, # apply to each column
                                FUN = function(x){
                                  temp <- c(mean(x), sd(x))  # find the mean and sd of each column and store as this vector
                                  names(temp) <- c("mean", "sd")  # apply names
                                  temp # return temp
                                }
)
custom_batting_summary
```

- We can also have additional arguments

```{r}
custom_batting_summary <- apply(X = my_batting |>
                                  select(where(is.numeric)),
                                MARGIN = 2, # apply to each column
                                FUN = function(x, trim){  # two arguments here
                                  temp <- c(mean(x, trim), sd(x))  # find the trimmed mean and sd of each column and store as this vector
                                  names(temp) <- c("mean", "sd")  # apply names
                                  temp # return temp
                                },
                                trim = 0.1 # define second argument here
)
custom_batting_summary
```

- Now lets try lapply()

- First create some list data:

```{r}
set.seed(10) # set randomization seed
my_list <- list(rnorm(100), runif(10), rgamma(40, shape = 1, rate =1))
my_list
```
- Find mean of each list element

```{r}
lapply(X = my_list, FUN = mean, trim = 0.1, na.rm = TRUE) # trim and na.rm get passed into mean as arguments
```

- Notes this returns a list of three elements, rather than a vector of the results

- sapply() tries to simplify if possible

```{r}
sapply(X = my_list, FUN = mean, trim = 0.1, na.rm = TRUE)
```

