---
title: "Week #9"
format: pdf
editor: visual
---

```{r}
library(tidyverse)
library(tidymodels)
```

## Modeling Concepts

```{r}
# read in bike data and create some variables
bike_data <- read_csv("https://www4.stat.ncsu.edu/~online/datasets/bikeDetails.csv")
bike_data <- bike_data |> 
  mutate(log_selling_price = log(selling_price), 
         log_km_driven = log(km_driven)) |>
  select(log_km_driven, log_selling_price, everything())
bike_data
# log selling price is the target(response) variable
```

- perform a regression task, i.e. predicting the response values

  - log_selling_price is the response variable

  - log_km_driven is the predictor variable

```{r}
ggplot(bike_data, aes(x = log_km_driven, y = log_selling_price)) +
  geom_point() +
  geom_smooth(method = "lm") # fits simple linear regression line
```

```{r}
ggplot(bike_data, aes(x = log_km_driven, y = log_selling_price)) +
  geom_point() +
  stat_smooth(method = "lm", formula = y ~ x + I(x^2))
# quadratic predictor (i.e. linear model, but the predictor is squared)
```

```{r}
ggplot(bike_data, aes(x = log_km_driven, y = log_selling_price)) +
  geom_point() +
  stat_smooth(method = "lm", formula = y ~ x + I(x^2) + I(x^3))
# here we have an x^3 term
```

- Need to quantify how well the model is predicting

```{r}
ggplot(bike_data, aes(x = log_km_driven, y = log_selling_price)) +
  geom_point() +
  geom_smooth()
# generalized additive model
```

- Also consider regression tree models (disjointed non-linear)

- Loss function: criteria used to fit a model (i.e. y-yhat, (y-yhat)\^2, or \|y-yhat\|)

- Goal is to optimize/minimize the loss function

- `lm()` is the linear model function

- use formula notation: `y ~ x -> response ~ model terms`

```{r}
slr_fit <- lm(log_selling_price ~ log_km_driven, data = bike_data)
slr_fit
```

- use summary() on the model to get p-values for the coefficients

```{r}
summary(slr_fit)
```

- use plot() on the model for fitted diagnostics

```{r}
plot(slr_fit)
```

- use `predict()` to test our fit with new data

```{r}
predict(slr_fit, newdata = data.frame(log_km_driven = c(log(1000), log(10000), log(100000))))

```

```{r}
# use exp to exponentiate (reverse) the log
exp(predict(slr_fit, newdata = data.frame(log_km_driven = c(log(1000), log(10000), log(100000)))))
```

- we use a [loss]{.underline} function to fit the model, and a [metric]{.underline} to evaluate the model

- metric can be the same as the loss, but not necessarily

- often we use RMSE (root mean squared error) as our metric

- Alternatives: MSE (mean squared error) or MAE (mean absolute error) or Huber Loss (doesn't penalize large misakes as much)

```{r}
# predictions for the x's the model was trained on
head(predict(slr_fit))
```

```{r}
# MSE
mean((bike_data$log_selling_price-predict(slr_fit))^2)
# RMSE
sqrt(mean((bike_data$log_selling_price-predict(slr_fit))^2))
```

- Ideally train the model on a training data set, and then test the model on a test data set.

```{r}
library(tidymodels)
set.seed(10)
bike_split <- initial_split(bike_data, prop = 0.7) # define function to perform 70/30 split
bike_train <- training(bike_split)
bike_test <- testing(bike_split)
bike_train
bike_test
```

- now create 3 different models with training data

```{r}
reg1 <- lm(log_selling_price ~ year, data = bike_train) # linear model based on year
coef(reg1)
```

```{r}
reg2 <- lm(log_selling_price ~ log_km_driven, data = bike_train) # linear model based on log km driven
coef(reg2)
```

```{r}
reg3 <- lm(log_selling_price ~ year + log_km_driven, data = bike_train) # linear model based on year and log km driven
coef(reg3)
```

- check predictions with first model on the training data

```{r}
bike_train |>
  mutate(mod1_preds = predict(reg1)) |>
  select(log_selling_price, year, mod1_preds)
```

```{r}
# our own calculation for training RMSE
sqrt(mean((bike_train$log_selling_price - predict(reg1))^2))
```

- alternatively there is rmse_vec() to do this

```{r}
#using yardstick on all three models
rmse_vec(bike_train$log_selling_price, predict(reg1))
rmse_vec(bike_train$log_selling_price, predict(reg2))
rmse_vec(bike_train$log_selling_price, predict(reg3))
```

- but really we should be testing on the test dataset

```{r}
#obtain the test set RMSE for each model
rmse_vec(bike_test$log_selling_price, predict(reg1, newdata = bike_test))
rmse_vec(bike_test$log_selling_price, predict(reg2, newdata = bike_test))
rmse_vec(bike_test$log_selling_price, predict(reg3, newdata = bike_test))
```

- so the third model has the lowest RMSE on the test data set

- alternatively, rmse() can be used instead of rmse_Vec() if we create the predictions within the dataframe

```{r}
# calculate rmse again with the first model on the test data set
bike_test |>
  select(log_selling_price, log_km_driven) |>
  mutate(model_1_preds = predict(reg1, newdata = bike_test)) |>
  rmse(truth = log_selling_price, estimate = model_1_preds)
```

- 10 Fold Cross-Validation

```{r}
nrow(bike_data) # confirm number of observations is 1061
```

```{r}
size_fold <- floor(nrow(bike_data)/10) # each fold will be 106 (10 folds)
size_fold
```
- Use sample() to reorder the 1061 indices and save as random_indices

```{r}
set.seed(8)
random_indices <- sample(1:nrow(bike_data), size = nrow(bike_data), replace = FALSE)
#see the random reordering
head(random_indices)
```

- Save folds in a list

```{r}
#create a list to save our folds in
folds <- list()
#now cycle through our random indices vector and take the appropriate observations to each fold
for(i in 1:10){
  if (i < 10) {
    fold_index <- seq(from = (i-1)*size_fold +1, to = i*size_fold, by = 1) # fold_index is 1:106, then 107:212, etc...
    folds[[i]] <- bike_data[random_indices[fold_index], ] # use fold_index and random indices to select from bike_data
  } else { #for i = 10 we need to have 1 more observation 
    fold_index <- seq(from = (i-1)*size_fold +1, to = length(random_indices), by = 1)
    folds[[i]] <- bike_data[random_indices[fold_index], ]
  }
}
folds[[1]]
```


- Now make a function to take in data and return with a list containing the number of folds of the data

```{r}
get_cv_splits <- function(data, num_folds){
  #get fold size
  size_fold <- floor(nrow(data)/num_folds)
  #get random indices to subset the data with
  random_indices <- sample(1:nrow(data), size = nrow(data), replace = FALSE)
  #create a list to save our folds in
  folds <- list()
  #now cycle through our random indices vector and take the appropriate observations to each fold
  for(i in 1:num_folds){
    if (i < num_folds) {
      fold_index <- seq(from = (i-1)*size_fold +1, to = i*size_fold, by = 1)
      folds[[i]] <- data[random_indices[fold_index], ]
    } else {
      fold_index <- seq(from = (i-1)*size_fold +1, to = length(random_indices), by = 1)
      folds[[i]] <- data[random_indices[fold_index], ]
    }
  }
  return(folds)
}
folds <- get_cv_splits(bike_data, 10)
```

- Now fit the data on 9 of 10 folds, while pulling out one as the test fold

```{r}
#set the test fold number
test_number <- 10
#pull out the testing fold
test <- folds[[test_number]]

#remove the test fold from the list and use the rest as training data
train_list <- folds[-test_number]
#note that train is only length 9
length(train_list)
```
- Condense the 9 folds into a single training set using reduce() and rbind()

```{r}
#combine the other folds into train set
train <- purrr::reduce(train_list, rbind)
#the other nine folds are now in train
train
```

- Write a function to fit the model, with x as a vector of predictors and y as a string for the response, and x as a vector of predictors

```{r}
find_test_metric <- function(y, x, folds, test_number){
  #pull out the testing fold
  test <- folds[[test_number]]
  #remove the test fold from the list
  train_list <- folds[-test_number]
  #combine the other folds into train set
  train <- purrr::reduce(train_list, rbind)
  #fit our model. reformulate allows us to take strings and turn that into a formula
  fit <- lm(reformulate(termlabels = x, response = y), data = train)
  #get test error
  (yardstick::rmse_vec(test[[y]], predict(fit, newdata = test)))^2
}

#double check it works
find_test_metric(y = "log_selling_price", 
                 x = "log_km_driven", 
                 folds = folds, 
                 test_number = 10)
```

- Now we can use map() to run through this 10 times, each using a different fold as the test

```{r}
folds <- get_cv_splits(bike_data, 10)
purrr::map(1:10, find_test_metric, 
           y = "log_selling_price", 
           x = "log_km_driven", 
           folds = folds)
```
- Combine all these values into one:  the RMSE of the average MSE

```{r}
sum_mse <- purrr::map(1:10, 
                      find_test_metric, 
                      y = "log_selling_price", 
                      x = "log_km_driven", 
                      folds = folds) |>
  reduce(.f = sum)
sqrt(sum_mse/10)
```

- Now put that into a function:

```{r}
find_cv <- function(y, x, folds, num_folds){
  sum_mse <- purrr::map(1:num_folds, 
                        find_test_metric, 
                        y = y, 
                        x =x, 
                        folds = folds) |>
    reduce(.f = sum)
  return(sqrt(sum_mse/num_folds))
}
find_cv("log_selling_price", "log_km_driven", folds, num_folds = 10)
```
- Let's compare different models on this dataset with 10 fold CV

```{r}
folds <- get_cv_splits(bike_data, 10)

cv_reg_1 <- find_cv("log_selling_price", "year", folds, num_folds = 10)
cv_reg_2 <- find_cv("log_selling_price", "log_km_driven", folds, num_folds = 10)
cv_reg_3 <- find_cv("log_selling_price", c("year", "log_km_driven"), folds, num_folds = 10)

c("year" = cv_reg_1, "log_km_driven" = cv_reg_2, "both" = cv_reg_3)
```

- Note, the model that uses both predictors has the lowest CV RMSE

- Fit this 'best model' using the full dataset

```{r}
best_fit <- lm(log_selling_price ~ year + log_km_driven, data = bike_data)
best_fit
```

