---
title: "Week #6"
format: pdf
editor: visual
---

## apply() family of functions

- apply() family of functions are faster than for loops
- apply() will apply a function over an array or matrix
- lapply() will apply a function to every element of a list
- sapply() is a simplified version

```{r}
library(Lahman)
library(dplyr)
library(tidyr)
```

```{r}
my_batting <- Batting[, c("playerID", "teamID", "G", "AB", "R", "H", "X2B", "X3B", "HR")] |>
  as_tibble()
my_batting
```

- Use apply() to find summaries of the columns

```{r}
apply(X = my_batting,
  MARGIN = 2, # apply to each column
  FUN = summary,
  na.rm = TRUE)
```

- Note: Because apply() assumes similar data types, summary gives us unhelpful results since everything is coerced to characters.

- Lets try again, this time just selecting numeric data

```{r}
apply(X = my_batting |>
        select(where(is.numeric)),
  MARGIN = 2, # apply to each column
  FUN = summary,
  na.rm = TRUE)
```

- Now we get useful results

- Sometimes we can define a temporary function with apply(). This is known as an *anonymous* function or *lambda* function

```{r}
custom_batting_summary <- apply(X = my_batting |>
                                  select(where(is.numeric)),
                                MARGIN = 2, # apply to each column
                                FUN = function(x){
                                  temp <- c(mean(x), sd(x))  # find the mean and sd of each column and store as this vector
                                  names(temp) <- c("mean", "sd")  # apply names
                                  temp # return temp
                                }
)
custom_batting_summary
```

- We can also have additional arguments

```{r}
custom_batting_summary <- apply(X = my_batting |>
                                  select(where(is.numeric)),
                                MARGIN = 2, # apply to each column
                                FUN = function(x, trim){  # two arguments here
                                  temp <- c(mean(x, trim), sd(x))  # find the trimmed mean and sd of each column and store as this vector
                                  names(temp) <- c("mean", "sd")  # apply names
                                  temp # return temp
                                },
                                trim = 0.1 # define second argument here
)
custom_batting_summary
```

- Now lets try lapply()

- First create some list data:

```{r}
set.seed(10) # set randomization seed
my_list <- list(rnorm(100), runif(10), rgamma(40, shape = 1, rate =1))
my_list
```
- Find mean of each list element

```{r}
lapply(X = my_list, FUN = mean, trim = 0.1, na.rm = TRUE) # trim and na.rm get passed into mean as arguments
```

- Notes this returns a list of three elements, rather than a vector of the results

- sapply() tries to simplify if possible

```{r}
sapply(X = my_list, FUN = mean, trim = 0.1, na.rm = TRUE)
```

## purrr() and List Columns

```{r}
library(purrr)
```

- purrr() is an alternative to apply() functions

- map() is similar to lapply() and always returns a list

```{r}
set.seed(10)
my_list <- list(rnorm(100), runif(10), rgamma(40, shape = 1, rate = 1))
map(my_list, mean) # select data, then map the mean function to each element of the list
```

```{r}
map(my_list, 2) # graps the second element of each list
```

- making anonymous functions with map()

```{r}
map(my_list, \(x) mean(x)) # in this case the custom function is just the mean
```

```{r}
map(my_list, \(x) max(x)-min(x)) # here the custom function is max-min
```

- there are versions of map functions to return vectors of specific type: map_dbl() for double, map_chr() for character, map_lgl() for logical

```{r}
map_dbl(my_list, mean) # returns result as a vector of doubles
```
- map2() can be used to apply a function to two similar lists

```{r}
my_list_2 <- list(rnorm(100), runif(10), rgamma(40, shape = 1, rate = 1))
map2(my_list, my_list_2, \(x, y) mean(x)-mean(y))
```

- pmap() can be used for any number of similar lists

```{r}
my_list_3 <- list(rnorm(100), runif(10), rgamma(40, shape = 1, rate = 1))

pmap(list(my_list, my_list_2, my_list_3), # send three lists into pmap
     \(x, y, z) (mean(x)-mean(y))/mean(z))
```

- walk() allows you to use a side-effect function (i.e. histogram, summary table, etc) but then continue operating on the original data

```{r}
par(mfrow = c(1, 3))
#now apply the function but still have the original data
my_list |>
  walk(hist) |>
  map_dbl(mean)
# without walk, the mean would be applied to the histogram object
```

- There are tons of purrr functions for mapping with various inputs/outputs

- List columns:  One column of data frame could be a list of multiple items

```{r}
irisWdiffs <- iris |>
                  as_tibble() |>
                  mutate(diffs = pmap(list(Sepal.Length, Sepal.Width, Petal.Length, Petal.Width), 
                                      \(x, y, z, w) list(x-y, x-z, x-w))) |> # create diffs, which calculates three items for each row:  x-y, x-z, x-w
                  select(diffs, everything())

irisWdiffs
```

- Now there are many options to grab the elements from this new diff column using pull() and/or pluck():

```{r}
irisWdiffs$diffs[[30]][[2]] # gets the x-z term of the 30th row from diffs
pull(irisWdiffs, diffs)[[30]][[2]] # this also works
pluck(irisWdiffs$diffs, 30, 2) # this also works
irisWdiffs |> pull(diffs) |> pluck(30, 2) # inside pipe chain
```

## Advanced Function Writing

- `...` for unnamed arguments

```{r}
standardize <- function(vector, center = TRUE, scale = TRUE, ...) { # here we allow for unnamed arguments that will be passed into mean and stdev below
    mean <- mean(vector, ...)
    stdev <- sd(vector, ...)
    if (center) {
        vector <- vector - mean
    }
    if (scale) {
        vector <- vector / stdev
    }
    return(list(result = vector, mean = mean, sd = stdev))
}
```
 
- we can use for things like na.rm

```{r}
standard_Ozone <- standardize(airquality$Ozone, na.rm = TRUE)
standard_Ozone$mean
standard_Ozone$sd
```

- another example showing how we can return/access the unnamed arguments

```{r}
f <- function(x, ...){
    unnamed <- list(...) # saving the unnamed arguments as a list
    modifyX <- x^2  
    return(list(newX = modifyX, elipses = unnamed))
}
f(x = 10, a = 1, b = list(char = "hey there", num = 1:3))
```

- tidyverse style functions

- example:

```{r}
iris |>
  group_by(Species) |>
  summarize(across(where(is.numeric), 
                   list("mean" = mean), 
                   .names = "{.fn}_{.col}"))
```

- Now create a function to do the same

```{r}
find_group_mean <- function(.df, group){
    .df |>
    group_by({{group}}) |> # {{}} is used to pass in unquoted column names to a tidyverse function
    summarize(across(where(is.numeric), 
                     list("mean" = mean), 
                     .names = "{.fn}_{.col}"))
}
find_group_mean(iris, Species)
```

- instead of {{}}, you can also do enquo() with !!() or quos() with !!!()

- those must be used with ... arguments

```{r}
find_group_mean <- function(.df, ...){
  group_vars <- quos(...) # puts the additional arguments as quotes
  .df |>
    group_by(!!!group_vars) |> # strips the quotes to pass the names in
    summarize(across(where(is.numeric), 
                     list("mean" = mean), 
                     .names = "{.fn}_{.col}"))
}
find_group_mean(CO2, Type, Treatment)
```

- Now lets try using inputs to name variables

```{r}
find_group_mean <- function(.df, group, column){
  group_name <- enquo(group)
  column_name <- enquo(column)
  column_label <- paste0("mean_", as_label(column_name))
  .df |>
    group_by(!!group_name) |>
    summarize(!!(column_label) := mean(!!column_name))
}
find_group_mean(iris, Species, Sepal.Length)
```

- Pipeable functions

  - transformation functions return a modification of the argument
  
  - side-effects don't (i.e. print() or write_csv()
  
- side-effect functions need invisible() to remain pipeable

```{r}
print_num_obs <- function(.df) {
    cat("The number of observations in the data set is ", 
        nrow(.df), 
        "\n", 
        sep = "")
  invisible(.df) # makes sure the .df continues down the pipe, but does not print
}
iris |> 
  print_num_obs() |>
  summarize(mean = mean(Sepal.Length))
```

- Lazy Evaluation: 

  - R normally evaluates arguments only when needed, and ignores the rest
  
  - Evaluation can be forced with force()
  
- Environments and Lexical Scoping

  - R first searches global environment for the function to be used, then subsequent packages that are loaded
  
  - Can see this path with search()
  
```{r}
search()
```

  - functions create temporary environments, which is why variables in functions don't conflict with variables/functions outside
  
  - functions will first look for variables within the temporary environment for the variable name, before looking outside
  
