---
title: "Week #9"
format: pdf
editor: visual
---

```{r}
library(tidyverse)
library(tidymodels)
library(glmnet)
```

## LASSO Models

- First CV with basic MLR and minimizing RMSE

```{r}
set.seed(10)
bike_data <- read_csv("https://www4.stat.ncsu.edu/~online/datasets/bikeDetails.csv")
bike_data <- bike_data |> 
  mutate(log_selling_price = log(selling_price), 
         log_km_driven = log(km_driven),
         owners = ifelse(owner == "1st owner", "single", "multiple")) |>
  select(log_km_driven, log_selling_price, everything())
#use tidymodel functions for splitting the data
bike_split <- initial_split(bike_data, prop = 0.7)
bike_train <- training(bike_split)
bike_test <- testing(bike_split)
```
```{r}
#create folds
bike_CV_folds <- vfold_cv(bike_train, 10)
```

```{r}
#set up how we'll fit our linear model
MLR_spec <- linear_reg() |>
  set_engine("lm")
```

```{r}
#define our MLR models
MLR_recipe1 <- recipe(log_selling_price ~ log_km_driven + owners + year, 
                      data = bike_train) |>
  step_dummy(owners)
MLR_recipe2 <- recipe(log_selling_price ~ log_km_driven + owners,
                      data = bike_train) |>
  step_dummy(owners) |>
  step_interact(~log_km_driven:starts_with("owner"))

MLR_recipe3 <- recipe(log_selling_price ~ log_km_driven + owners + year,
                      data = bike_train) |>
  step_dummy(owners) |>
  step_interact(~log_km_driven:starts_with("owner") + log_km_driven:year + starts_with("owner"):year)
```

```{r}
MLR_wkf1 <- workflow() |>
  add_recipe(MLR_recipe1) |>
  add_model(MLR_spec)

MLR_wkf2 <- workflow() |>
  add_recipe(MLR_recipe2) |>
  add_model(MLR_spec)

MLR_wkf3 <- workflow() |>
  add_recipe(MLR_recipe3) |>
  add_model(MLR_spec)
```

```{r}
MLR_fit1 <-  MLR_wkf1 |>
  fit_resamples(bike_CV_folds)

MLR_fit2 <- MLR_wkf2 |>
  fit_resamples(bike_CV_folds) 

MLR_fit3 <- MLR_wkf3 |>
  fit_resamples(bike_CV_folds)
```

```{r}
rbind(MLR_fit1 |> collect_metrics() |> filter(.metric == "rmse"),
      MLR_fit2 |> collect_metrics() |> filter(.metric == "rmse"),
      MLR_fit3 |> collect_metrics() |> filter(.metric == "rmse")) |> 
  mutate(Model = c("Model 1", "Model 2", "Model 3")) |>
  select(Model, mean, n, std_err)
```
```{r}
MLR_final <-  MLR_wkf3 |>
  fit(bike_train)
tidy(MLR_final)
```

- Now fit a LASSO Model using CV

```{r}
#set up how we'll fit our LASSO model
#code modified from https://juliasilge.com/blog/lasso-the-office/
LASSO_recipe <- recipe(log_selling_price ~ log_km_driven + owners + year, 
                      data = bike_train) |>
  step_dummy(owners) |>
  step_normalize(log_km_driven, year)
```

```{r}
LASSO_spec <- linear_reg(penalty = tune(), mixture = 1) |>
  set_engine("glmnet")
```

```{r}
LASSO_wkf <- workflow() |>
  add_recipe(LASSO_recipe) |>
  add_model(LASSO_spec)
LASSO_wkf
```

```{r}
#A warning will occur for one value of the tuning parameter, safe to ignore
LASSO_grid <- LASSO_wkf |>
  tune_grid(resamples = bike_CV_folds,
            grid = grid_regular(penalty(), levels = 200),
            save_workflow = TRUE) 
LASSO_grid
```

```{r}
LASSO_grid |>
  collect_metrics() |>
  filter(.metric == "rmse")
```

```{r}
LASSO_grid |>
  collect_metrics() |>
  filter(.metric == "rmse") |>
  ggplot(aes(penalty, mean, color = .metric)) +
  geom_line()
```

- so RMSE dips very slightly at first, then increases with the penalty until it reaches a maximum (where all main effects and interaction coefficients are zero and we're left with a flat line at the mean)

```{r}
lowest_rmse <- LASSO_grid |>
  select_best(metric = "rmse")
lowest_rmse
```

```{r}
LASSO_wkf |>
  finalize_workflow(lowest_rmse)
```

```{r}
#fit it to the entire training set to see the model fit
LASSO_final <- LASSO_wkf |>
  finalize_workflow(lowest_rmse) |>
  fit(bike_train)
tidy(LASSO_final)
```

- Two ways to see results of model applied to test data:

```{r}
MLR_wkf3 |>
  last_fit(bike_split) |>
  collect_metrics()
LASSO_wkf |>
  finalize_workflow(lowest_rmse) |>
  last_fit(bike_split) |>
  collect_metrics()

```

```{r}
MLR_final |>
  predict(bike_test) |>
  pull() |>
  rmse_vec(truth = bike_test$log_selling_price)

LASSO_final |>
  predict(bike_test) |>
  pull() |>
  rmse_vec(truth = bike_test$log_selling_price)
```

- So actually the MLR model is better than the LASSO model when used vs the test set

- Last step is to refit the best model on the full data set for future use

```{r}
final_model <- MLR_wkf3 |>
  fit(bike_data) 
tidy(final_model)
```

```{r}
almost_usual_fit <- extract_fit_parsnip(final_model)
usual_fit <- almost_usual_fit$fit
summary(usual_fit)
```

## Logistic Regression Models

- Logistic regression useful when we have a binary response variable

- Attempting to do linear regression with binary response doesn't work.  Here's an example:

```{r}
library(tidyverse)
heart_data <- read_csv("https://www4.stat.ncsu.edu/online/datasets/heart.csv") |>
  filter(RestingBP > 0) #remove one value
heart_data |> select(HeartDisease, everything()) #Cholesterol has many values set to 0 so we ignore that
```

```{r}
heart_data |>
  group_by(HeartDisease) |> 
  summarize(count = n())
```

```{r}
heart_data |>
  group_by(HeartDisease) |>
  summarize(mean_Age = mean(Age),
            mean_RestingBP  = mean(RestingBP))
```

```{r}
ggplot(heart_data, aes(x = Age, y = HeartDisease, color = RestingBP)) +
         geom_point() +
  geom_smooth(method = "lm")
```

```{r}
ggplot(heart_data, aes(x = Age, y = HeartDisease, color = RestingBP)) +
         geom_jitter() + # jitter helps to spread out overlapping points
  geom_smooth(method = "lm")
```

```{r}
# show heart disease rates by age groups
Age_x <- seq(from = min(heart_data$Age), to = max(heart_data$Age), length = 20)
heart_data_grouped <- heart_data |>
  mutate(Age_groups = cut(Age, breaks = Age_x)) |>
  group_by(Age_groups) |>
  summarize(HeartDisease_mean = mean(HeartDisease), counts = n())
heart_data_grouped
```

```{r}
ggplot(data = heart_data, aes(x = Age, y = HeartDisease)) +
  geom_jitter(aes(color = RestingBP)) +
  geom_point(data = heart_data_grouped, aes(x = Age_x, y = HeartDisease_mean, size = counts)) +
  geom_smooth(method = "lm", color = "Green")
# now the line makes a little more sense with these groups shown on the graph
```
- Note, line fits will have regions where resonse (proportion) is less than 0 or greater than 1!

- Let's do a logistic regression model fit

```{r}
library(tidymodels)
set.seed(3557)
heart_data <- heart_data |> mutate(HeartDisease = factor(HeartDisease)) # make sure response is made into a factor
heart_split <- initial_split(heart_data, prop = 0.8)
heart_train <- training(heart_split)
heart_test <- testing(heart_split)
heart_CV_folds <- vfold_cv(heart_train, 10)
```

```{r}
# two predictors
LR1_rec <- recipe(HeartDisease ~ Age + Sex, 
                  data = heart_train) |>
  step_normalize(Age) |>
  step_dummy(Sex)

# five predictors
LR2_rec <- recipe(HeartDisease ~ Age + Sex + ChestPainType + RestingBP + RestingECG, 
                  data = heart_train) |>
  step_normalize(all_numeric(), -HeartDisease) |>
  step_dummy(Sex, ChestPainType, RestingECG)

# seven predictors
LR3_rec <- recipe(HeartDisease ~ Age + Sex + ChestPainType + RestingBP + RestingECG + MaxHR + ExerciseAngina, 
                  data = heart_train) |>
  step_normalize(all_numeric(), -HeartDisease) |>
  step_dummy(Sex, ChestPainType, RestingECG, ExerciseAngina)

# this is just to check what we've got
LR3_rec |> prep(heart_train) |> bake(heart_train) |> colnames()

# numeric variables are normalized
```

```{r}
# selecting logistic regression model
LR_spec <- logistic_reg() |>
  set_engine("glm")
```

```{r}
LR1_wkf <- workflow() |>
  add_recipe(LR1_rec) |>
  add_model(LR_spec)
LR2_wkf <- workflow() |>
  add_recipe(LR2_rec) |>
  add_model(LR_spec)
LR3_wkf <- workflow() |>
  add_recipe(LR3_rec) |>
  add_model(LR_spec)
```

- Instead of evaluating model based on RMSE, we can do accuracy (correct predictions/ total predictions) or Log-loss

```{r}
# fit by minimizing log-loss
LR1_fit <- LR1_wkf |>
  fit_resamples(heart_CV_folds, metrics = metric_set(accuracy, mn_log_loss))
LR2_fit <- LR2_wkf |>
  fit_resamples(heart_CV_folds, metrics = metric_set(accuracy, mn_log_loss))
LR3_fit <- LR3_wkf |>
  fit_resamples(heart_CV_folds, metrics = metric_set(accuracy, mn_log_loss))
```

```{r}
# collect metrics
rbind(LR1_fit |> collect_metrics(),
      LR2_fit |> collect_metrics(),
      LR3_fit |> collect_metrics()) |>
  mutate(Model = c("Model1", "Model1", "Model2", "Model2", "Model3", "Model3")) |>
  select(Model, everything())
```

- now take our best model (model 3 by lowest log-loss and highest accuracy)

```{r}
#compare to proportion of 1's in training data
mean(heart_train$HeartDisease == "1")

# sanity check, accuracy should at least beat .56
```
- Confusion Matrix on third workflow gives us breakdown of the correct and incorrect predictions

```{r}
LR_train_fit <- LR3_wkf |>
  fit(heart_train)
conf_mat(heart_train |> mutate(estimate = LR_train_fit |> predict(heart_train) |> pull()), #data
         HeartDisease, #truth
         estimate) #estimate from model
```

- Now evaluate our model on the test set: metrics and confusion matrix
```{r}
LR3_wkf |>
  last_fit(heart_split, metrics = metric_set(accuracy, mn_log_loss)) |>
  collect_metrics()

conf_mat(heart_test |> mutate(estimate = LR_train_fit |> predict(heart_test) |> pull()), HeartDisease, estimate)
```
- Do a final fit of the favored model using the full data set

```{r}
final_model <- LR3_wkf |>
  fit(heart_data)
tidy(final_model)
```

