---
title: "Week #9"
format: pdf
editor: visual
---

```{r}
library(tidyverse)
library(tidymodels)
library(glmnet)
```

## LASSO Models

- First CV with basic MLR and minimizing RMSE

```{r}
set.seed(10)
bike_data <- read_csv("https://www4.stat.ncsu.edu/~online/datasets/bikeDetails.csv")
bike_data <- bike_data |> 
  mutate(log_selling_price = log(selling_price), 
         log_km_driven = log(km_driven),
         owners = ifelse(owner == "1st owner", "single", "multiple")) |>
  select(log_km_driven, log_selling_price, everything())
#use tidymodel functions for splitting the data
bike_split <- initial_split(bike_data, prop = 0.7)
bike_train <- training(bike_split)
bike_test <- testing(bike_split)
```
```{r}
#create folds
bike_CV_folds <- vfold_cv(bike_train, 10)
```

```{r}
#set up how we'll fit our linear model
MLR_spec <- linear_reg() |>
  set_engine("lm")
```

```{r}
#define our MLR models
MLR_recipe1 <- recipe(log_selling_price ~ log_km_driven + owners + year, 
                      data = bike_train) |>
  step_dummy(owners)
MLR_recipe2 <- recipe(log_selling_price ~ log_km_driven + owners,
                      data = bike_train) |>
  step_dummy(owners) |>
  step_interact(~log_km_driven:starts_with("owner"))

MLR_recipe3 <- recipe(log_selling_price ~ log_km_driven + owners + year,
                      data = bike_train) |>
  step_dummy(owners) |>
  step_interact(~log_km_driven:starts_with("owner") + log_km_driven:year + starts_with("owner"):year)
```

```{r}
MLR_wkf1 <- workflow() |>
  add_recipe(MLR_recipe1) |>
  add_model(MLR_spec)

MLR_wkf2 <- workflow() |>
  add_recipe(MLR_recipe2) |>
  add_model(MLR_spec)

MLR_wkf3 <- workflow() |>
  add_recipe(MLR_recipe3) |>
  add_model(MLR_spec)
```

```{r}
MLR_fit1 <-  MLR_wkf1 |>
  fit_resamples(bike_CV_folds)

MLR_fit2 <- MLR_wkf2 |>
  fit_resamples(bike_CV_folds) 

MLR_fit3 <- MLR_wkf3 |>
  fit_resamples(bike_CV_folds)
```

```{r}
rbind(MLR_fit1 |> collect_metrics() |> filter(.metric == "rmse"),
      MLR_fit2 |> collect_metrics() |> filter(.metric == "rmse"),
      MLR_fit3 |> collect_metrics() |> filter(.metric == "rmse")) |> 
  mutate(Model = c("Model 1", "Model 2", "Model 3")) |>
  select(Model, mean, n, std_err)
```
```{r}
MLR_final <-  MLR_wkf3 |>
  fit(bike_train)
tidy(MLR_final)
```

- Now fit a LASSO Model using CV

```{r}
#set up how we'll fit our LASSO model
#code modified from https://juliasilge.com/blog/lasso-the-office/
LASSO_recipe <- recipe(log_selling_price ~ log_km_driven + owners + year, 
                      data = bike_train) |>
  step_dummy(owners) |>
  step_normalize(log_km_driven, year)
```

```{r}
LASSO_spec <- linear_reg(penalty = tune(), mixture = 1) |>
  set_engine("glmnet")
```

```{r}
LASSO_wkf <- workflow() |>
  add_recipe(LASSO_recipe) |>
  add_model(LASSO_spec)
LASSO_wkf
```

```{r}
#A warning will occur for one value of the tuning parameter, safe to ignore
LASSO_grid <- LASSO_wkf |>
  tune_grid(resamples = bike_CV_folds,
            grid = grid_regular(penalty(), levels = 200),
            save_workflow = TRUE) 
LASSO_grid
```

```{r}
LASSO_grid |>
  collect_metrics() |>
  filter(.metric == "rmse")
```

```{r}
LASSO_grid |>
  collect_metrics() |>
  filter(.metric == "rmse") |>
  ggplot(aes(penalty, mean, color = .metric)) +
  geom_line()
```

- so RMSE dips very slightly at first, then increases with the penalty until it reaches a maximum (where all main effects and interaction coefficients are zero and we're left with a flat line at the mean)

```{r}
lowest_rmse <- LASSO_grid |>
  select_best(metric = "rmse")
lowest_rmse
```

```{r}
LASSO_wkf |>
  finalize_workflow(lowest_rmse)
```

```{r}
#fit it to the entire training set to see the model fit
LASSO_final <- LASSO_wkf |>
  finalize_workflow(lowest_rmse) |>
  fit(bike_train)
tidy(LASSO_final)
```

- Two ways to see results of model applied to test data:

```{r}
MLR_wkf3 |>
  last_fit(bike_split) |>
  collect_metrics()
LASSO_wkf |>
  finalize_workflow(lowest_rmse) |>
  last_fit(bike_split) |>
  collect_metrics()

```

```{r}
MLR_final |>
  predict(bike_test) |>
  pull() |>
  rmse_vec(truth = bike_test$log_selling_price)

LASSO_final |>
  predict(bike_test) |>
  pull() |>
  rmse_vec(truth = bike_test$log_selling_price)
```

- So actually the MLR model is better than the LASSO model when used vs the test set

- Last step is to refit the best model on the full data set for future use

```{r}
final_model <- MLR_wkf3 |>
  fit(bike_data) 
tidy(final_model)
```

```{r}
almost_usual_fit <- extract_fit_parsnip(final_model)
usual_fit <- almost_usual_fit$fit
summary(usual_fit)
```

## Logistic Regression Models

- Logistic regression useful when we have a binary response variable

- Attempting to do linear regression with binary response doesn't work.  Here's an example:

```{r}
library(tidyverse)
heart_data <- read_csv("https://www4.stat.ncsu.edu/online/datasets/heart.csv") |>
  filter(RestingBP > 0) #remove one value
heart_data |> select(HeartDisease, everything()) #Cholesterol has many values set to 0 so we ignore that
```

```{r}
heart_data |>
  group_by(HeartDisease) |> 
  summarize(count = n())
```

```{r}
heart_data |>
  group_by(HeartDisease) |>
  summarize(mean_Age = mean(Age),
            mean_RestingBP  = mean(RestingBP))
```

```{r}
ggplot(heart_data, aes(x = Age, y = HeartDisease, color = RestingBP)) +
         geom_point() +
  geom_smooth(method = "lm")
```

```{r}
ggplot(heart_data, aes(x = Age, y = HeartDisease, color = RestingBP)) +
         geom_jitter() + # jitter helps to spread out overlapping points
  geom_smooth(method = "lm")
```

```{r}
# show heart disease rates by age groups
Age_x <- seq(from = min(heart_data$Age), to = max(heart_data$Age), length = 20)
heart_data_grouped <- heart_data |>
  mutate(Age_groups = cut(Age, breaks = Age_x)) |>
  group_by(Age_groups) |>
  summarize(HeartDisease_mean = mean(HeartDisease), counts = n())
heart_data_grouped
```

```{r}
ggplot(data = heart_data, aes(x = Age, y = HeartDisease)) +
  geom_jitter(aes(color = RestingBP)) +
  geom_point(data = heart_data_grouped, aes(x = Age_x, y = HeartDisease_mean, size = counts)) +
  geom_smooth(method = "lm", color = "Green")
# now the line makes a little more sense with these groups shown on the graph
```
- Note, line fits will have regions where resonse (proportion) is less than 0 or greater than 1!

- Let's do a logistic regression model fit

```{r}
library(tidymodels)
set.seed(3557)
heart_data <- heart_data |> mutate(HeartDisease = factor(HeartDisease)) # make sure response is made into a factor
heart_split <- initial_split(heart_data, prop = 0.8)
heart_train <- training(heart_split)
heart_test <- testing(heart_split)
heart_CV_folds <- vfold_cv(heart_train, 10)
```

```{r}
# two predictors
LR1_rec <- recipe(HeartDisease ~ Age + Sex, 
                  data = heart_train) |>
  step_normalize(Age) |>
  step_dummy(Sex)

# five predictors
LR2_rec <- recipe(HeartDisease ~ Age + Sex + ChestPainType + RestingBP + RestingECG, 
                  data = heart_train) |>
  step_normalize(all_numeric(), -HeartDisease) |>
  step_dummy(Sex, ChestPainType, RestingECG)

# seven predictors
LR3_rec <- recipe(HeartDisease ~ Age + Sex + ChestPainType + RestingBP + RestingECG + MaxHR + ExerciseAngina, 
                  data = heart_train) |>
  step_normalize(all_numeric(), -HeartDisease) |>
  step_dummy(Sex, ChestPainType, RestingECG, ExerciseAngina)

# this is just to check what we've got
LR3_rec |> prep(heart_train) |> bake(heart_train) |> colnames()

# numeric variables are normalized
```

```{r}
# selecting logistic regression model
LR_spec <- logistic_reg() |>
  set_engine("glm")
```

```{r}
LR1_wkf <- workflow() |>
  add_recipe(LR1_rec) |>
  add_model(LR_spec)
LR2_wkf <- workflow() |>
  add_recipe(LR2_rec) |>
  add_model(LR_spec)
LR3_wkf <- workflow() |>
  add_recipe(LR3_rec) |>
  add_model(LR_spec)
```

- Instead of evaluating model based on RMSE, we can do accuracy (correct predictions/ total predictions) or Log-loss

```{r}
# fit by minimizing log-loss
LR1_fit <- LR1_wkf |>
  fit_resamples(heart_CV_folds, metrics = metric_set(accuracy, mn_log_loss))
LR2_fit <- LR2_wkf |>
  fit_resamples(heart_CV_folds, metrics = metric_set(accuracy, mn_log_loss))
LR3_fit <- LR3_wkf |>
  fit_resamples(heart_CV_folds, metrics = metric_set(accuracy, mn_log_loss))
```

```{r}
# collect metrics
rbind(LR1_fit |> collect_metrics(),
      LR2_fit |> collect_metrics(),
      LR3_fit |> collect_metrics()) |>
  mutate(Model = c("Model1", "Model1", "Model2", "Model2", "Model3", "Model3")) |>
  select(Model, everything())
```

- now take our best model (model 3 by lowest log-loss and highest accuracy)

```{r}
#compare to proportion of 1's in training data
mean(heart_train$HeartDisease == "1")

# sanity check, accuracy should at least beat .56
```
- Confusion Matrix on third workflow gives us breakdown of the correct and incorrect predictions

```{r}
LR_train_fit <- LR3_wkf |>
  fit(heart_train)
conf_mat(heart_train |> mutate(estimate = LR_train_fit |> predict(heart_train) |> pull()), #data
         HeartDisease, #truth
         estimate) #estimate from model
```

- Now evaluate our model on the test set: metrics and confusion matrix
```{r}
LR3_wkf |>
  last_fit(heart_split, metrics = metric_set(accuracy, mn_log_loss)) |>
  collect_metrics()

conf_mat(heart_test |> mutate(estimate = LR_train_fit |> predict(heart_test) |> pull()), HeartDisease, estimate)
```
- Do a final fit of the favored model using the full data set

```{r}
final_model <- LR3_wkf |>
  fit(heart_data)
tidy(final_model)
```

## Tree Based Methods

```{r}
library(tree) #rpart is also often used
```

- Here is an example tree model

```{r}
fitTree <- tree(dist ~ speed, data = cars) #default splitting is deviance
plot(fitTree)
text(fitTree)
```

- Compare tree results (orange) to simple linear regression (blue)

```{r}
library(tidyverse)
ggplot(cars, aes(x = speed, y = dist)) + 
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, size = 2) + 
  geom_segment(x = 0, xend = 9.5, y = 10.67, yend = 10.67, col = "Orange", size = 2) +
  geom_segment(x = 9.5, xend = 12.5, y = 23.22, yend = 23.22, col = "Orange", size = 2) +
  geom_segment(x = 12.5, xend = 17.5, y = 39.75, yend = 39.75, col = "Orange", size = 2) +
  geom_segment(x = 17.5, xend = 23.5, y = 55.71, yend = 55.71, col = "Orange", size = 2) +
  geom_segment(x = 23.5, xend = max(cars$speed), y = 92, yend = 92, col = "Orange", size = 2)
```

- Optimal splits are determined based on minimizing sum of squared residuals

- Let's look at the bike data again and compare the regreesion tree method to the MLR and LASSO models

```{r}
set.seed(10)
library(tidyverse)
library(tidymodels)
bike_data <- read_csv("https://www4.stat.ncsu.edu/~online/datasets/bikeDetails.csv") |>
  mutate(log_selling_price = log(selling_price),
         log_km_driven = log(km_driven),
         owners = ifelse(owner == "1st owner", "single", "multiple")) |>
  select(-selling_price)
```
```{r}
#save creation of new variables for now!
bike_split <- initial_split(bike_data, prop = 0.7)
bike_train <- training(bike_split)
bike_test <- testing(bike_split)
bike_train
```

```{r}
bike_CV_folds <- vfold_cv(bike_train, 10)
```

```{r}
tree_rec <- recipe(log_selling_price ~ ., data = bike_train) |>
  update_role(name, new_role = "ID") |>
  step_log(km_driven) |>
  step_rm(ex_showroom_price) |>
  step_dummy(owner, seller_type) |>
  step_normalize(all_numeric(), -all_outcomes()) # Note:  normalization is NOT required for tree models, but we do it here for consistency with the MLR and LASSO recipes
tree_rec
```

- Note:  no need for interaction terms with the tree model.  The tree method inherently includes these relationships due to the splitting

```{r}
tree_mod <- decision_tree(tree_depth = tune(), # maximum number of splits from root to leaf
                          min_n = 20, # minimum number of observations to be eligible for splitting
                          cost_complexity = tune()) |> # 'prunes' the tree after it is built (i.e. penalizing complexity)
  set_engine("rpart") |>
  set_mode("regression")
```

```{r}
tree_wkf <- workflow() |>
  add_recipe(tree_rec) |>
  add_model(tree_mod)
```


```{r}
# without specifying the tuning options, they are selected randomly (10 total combinations)
temp <- tree_wkf |> 
  tune_grid(resamples = bike_CV_folds) #uses defaults from dials
temp |> 
  collect_metrics()
```

```{r}
# Here we say specifically how many combinations of tuning variables we want (10 x 5 = 50 combinations)
tree_grid <- grid_regular(cost_complexity(),
                          tree_depth(),
                          levels = c(10, 5))
```

```{r}
tree_fits <- tree_wkf |> 
  tune_grid(resamples = bike_CV_folds,
            grid = tree_grid)
tree_fits
```

```{r}
tree_fits |>
  collect_metrics()
```

- Let's plot these metrics as a function of the tuned model parameters:

```{r}
tree_fits |>
  collect_metrics() |>
  mutate(tree_depth = factor(tree_depth)) |>
  ggplot(aes(cost_complexity, mean, color = tree_depth)) +
  geom_line(size = 1.5, alpha = 0.6) +
  geom_point(size = 2) +
  facet_wrap(~ .metric, scales = "free", nrow = 2) +
  scale_x_log10(labels = scales::label_number()) +
  scale_color_viridis_d(option = "plasma", begin = .9, end = 0)
```

```{r}
# just look at smallest RMSE
tree_fits |>
  collect_metrics() |>
  filter(.metric == "rmse") |>
  arrange(mean)
```


- so the lowest RMSE tree model uses 1e-3 cost_complexity, tree_depth of 4, and min_n of 20

```{r}
# this selects the best model's tuning parameters

tree_best_params <- select_best(tree_fits, metric = "rmse")
tree_best_params
```

```{r}
tree_final_wkf <- tree_wkf |>
  finalize_workflow(tree_best_params)
```

```{r}
tree_final_fit <- tree_final_wkf |>
  last_fit(bike_split)
tree_final_fit
```

```{r}
# here is the RMSE of the selected model on the full training set vs the test set
tree_final_fit |>
  collect_metrics()
```

- Here's the details of this final model:

```{r}
tree_final_model <- extract_workflow(tree_final_fit) 
tree_final_model
```

- The above tree model is best viewed as a plot (here saved as a higher res .png)

```{r}
#| fig-width: 4
#| fig-height: 2
#| dpi: 300

library(rpart.plot)

tree_final_model |>
  extract_fit_engine() |>
  rpart.plot::rpart.plot(roundint = FALSE, digits = 4)


# Note: The owner_X2nd.owner is a dummy variable for "2nd owner", and the >= 1.2 can be ignored.  Just think of the split as 0/1
```

- Now we can compare this to the MLR and LASSO models

```{r}
#set up how we'll fit our linear model
MLR_spec <- linear_reg() |>
  set_engine("lm")
#define the model
MLR_recipe3 <- recipe(log_selling_price ~ log_km_driven + owners + year,
                      data = bike_train) |>
  step_dummy(owners) |>
  step_interact(~log_km_driven:starts_with("owner") + log_km_driven:year + starts_with("owner"):year)
#workflow setup
MLR_wkf3 <- workflow() |>
  add_recipe(MLR_recipe3) |>
  add_model(MLR_spec)
#fit model on full training data
MLR_final <-  MLR_wkf3 |>
  last_fit(bike_split)
```

```{r}
#set up how we'll fit our LASSO model
LASSO_spec <- linear_reg(penalty = tune(), mixture = 1) |>
  set_engine("glmnet")
#define the model
LASSO_recipe <- recipe(log_selling_price ~ log_km_driven + owners + year, 
                      data = bike_train) |>
  step_dummy(owners) |>
  step_normalize(log_km_driven, year)
#workflow setup
LASSO_wkf <- workflow() |>
  add_recipe(LASSO_recipe) |>
  add_model(LASSO_spec)
#find the best tuning parameter
LASSO_grid <- LASSO_wkf |>
  tune_grid(resamples = bike_CV_folds,
            grid = grid_regular(penalty(), levels = 200)) 
```

```{r}
#finalize the workflow and find the best model on the training set
LASSO_final <- LASSO_wkf |>
  finalize_workflow(LASSO_grid |>
                      select_best(metric = "rmse")) |>
  last_fit(bike_split)
```

```{r}
rbind(
  MLR_final |>
    collect_metrics() |> 
    mutate(Model = "MLR", .before = ".metric"), 
  LASSO_final |>
    collect_metrics() |> 
    mutate(Model = "LASSO", .before = ".metric"), 
  tree_final_fit |>
    collect_metrics() |> 
    mutate(Model = "TREE", .before = ".metric")
)
```

- So the MLR model had the lowest RMSE

- Finally, fit the best model (MLR) to the full data set

```{r}
final_model <- MLR_wkf3 |>
  fit(bike_data) 
tidy(final_model)
```

